Properties and Bayesian fitting of restricted Boltzmann machines

### 2017-11-27 writing/paper-rbm/chapters/rbm.Rmd
getting ready for resubmission

In addition to such degeneracy, we find that RBMs can easily exhibit types of instability, related to how sensitive the model probability structure can be to small changes in data outcomes. In practice, this may be seen when a single pixel change in an image results in a wildly different classification in an image classification problem. Occurrences of such behavior have recently been documented in RBMs [@li2014biclustering], as well as other deep architectures [@szegedy2013intriguing; @nguyen2014deep]. We describe potential issues of model instability, degeneracy and uninterpretability for the RBM class, which are related properties, and examine the presence of these in Section \@ref(explorations-of-model-properties-through-simulation) through simulations of small, manageable examples.

@@

Notions of weight-decay (penalization) and sparsity (regularization) have been suggested in the RBM literature as practical remedies for both over-fitting and poor mixing in the Markov chain during fitting [@hinton2010practical; @tieleman2008training; @cho2012tikhonov]. Both $L_1$ and $L_2$ type penalties are mentioned to achieve weight-decay, though the benefits of these particular forms are unknown in different situations. However, the degree to which regularization and penalization are used by practitioners is not clear because these concepts are not (by default) a part of the accepted fitting methodology [@hinton2002training; @carreira2005contrastive]. In this paper, we attempt to address the concerns with overfitting and poor mixing in an alternative, and perhaps more transparent manner, via specification of a Bayesian prior in Section \@ref(bayesian-model-fitting).

This paper is structured as follows. Section \@ref(background) formally defines the RBM including the joint distribution of hidden and visible variables and explains the model's connection to deep learning. Additionally, measures of model impropriety and methods of quantifying/detecting it are defined. Section \@ref(explorations-of-model-properties-through-simulation) details our explorations into the model behavior and potential propriety issues with the RBM class. We examine three Bayesian fitting techniques intended to avoid model impropriety issues raised in Section \@ref(model-fitting) and conclude with a discussion in Section \@ref(discussion). Appendix \@ref(appendix-rbm) provides proofs for results on RBM parameterizations and data codings described in Section \@ref(data-encoding).

While applications of the RBM have claimed some successes in classification problems, it is unclear if the model class allows one to go beyond fitting to other statistical matters of importance in using the models, such as quantification of the uncertainty of estimation and the formulation of predictive distributions. These are closely tied to the probability properties of the RBM model for explaining data generation and the current exposition intends to contribute to a better understanding in this direction.

@@

Considering exponential families of distributions, @schweinberger2011instability introduced a concept of model deficiency related to *instability*. Instability can be roughly thought of as excessive sensitivity in the model, where small changes in the components of potential data outcomes, $\boldsymbol x$, may lead to substantial changes in the probability mass function $f_{\boldsymbol \theta}(\boldsymbol x)$. Furthermore, model instability can be viewed on a spectrum of potential sensitivity in probability structure, with model degeneracy included as a extreme or limiting case of instability. To quantify "instability" more rigorously (particularly beyond the definition given by @schweinberger2011instability) it is useful to consider how RBM models might be expanded to incorporate more and more visibles. When increasing the size of RBM models, it becomes necessary to grow the number of model parameters (and in this process one may also arbitrarily expand the number of hidden variables used). To this end, let $\boldsymbol \theta_{\nv} \equiv (\theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}, \theta_{11}, \dots, \theta_{\nv \nh}), \nv \ge 1$, denote an element of a sequence of RBM parameters indexed by the number $\nv$ of visibles $(V_1, \dots, V_\nv)$ and define a (scaled) extremal log-probability ratio of the RBM model at $\boldsymbol \theta_{\nv}$ as
\begin{align}
\frac{1}{\nv} \log \left[\frac{\max\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{\min\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}\right] \equiv \frac{1}{\nv} \text{ELPR}(\boldsymbol \theta_\nv) (\#eq:elpr)
\end{align}
where $P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv) \propto \sum_{(h_1, \dots, h_\nh) \in \mathcal{C}^\nh}\exp\left(\sum_{i = 1}^\nv \theta_{v_i} v_i + \sum_{j = 1}^\nh \theta_{h_j} h_j + \sum_{i = 1}^\nv \sum_{j = 1}^\nh \theta_{ij} v_i h_j\right)$ is the RBM probability of observing outcome $(v_1, \dots, v_\nv)$ for the visible variables $(V_1, \dots, V_\nv)$ under parameter vector $\boldsymbol \theta_\nv$, after marginalization of hidden variables.

In formulating a RBM model for a potentially large number of visibles (i.e., as $\nv \rightarrow \infty$), we will say that the ratio (\@ref(eq:elpr)) needs to stay bounded for a sequence of RBM models to be stable. That is, we make the following convention.
```{definition, name = "S-unstable RBM", echo=TRUE}
Let $\boldsymbol \theta_\nv \in \mathbb{R}^{\nv + \nh +  \nh*\nv}, \nv \ge 1$, be an element of a sequence of RBM parameters where the number of hiddens, $\nh \equiv \nh(\nv) \ge 1$, can be an arbitrary function of the number $\nv$ of visibles. A RBM model formulation is \emph{Schweinberger-unstable} or \emph{S-unstable} if
\begin{align*}
\lim\limits_{\nv \rightarrow \infty} \frac{1}{\nv} \text{ELPR}(\boldsymbol \theta_\nv) = \infty.
\end{align*}
```
In other words, a RBM model sequence is unstable if, given any $c > 0$, there exists an integer $n_c > 0$ so that $\frac{1}{\nv}\text{ELPR}(\boldsymbol \theta_\nv) > c$ for all $\nv \ge n_c$. This definition of *S-unstable* is a generalization or re-interpretation of the "unstable" concept of @schweinberger2011instability in that here RBM models for visibles $(v_1, \dots, v_\nv)$ do not form an exponential family and the dimensionality of $\boldsymbol \theta_\nv$ is not fixed, but rather grows with $\nv$.

@@

```{proposition, label = "instab1", echo=TRUE}
Let $c > 0$ and let $\text{ELPR}(\boldsymbol \theta_\nv)$ be as in (\@ref(eq:elpr)) for an integer $\nv \ge 1$. If $\frac{1}{\nv}\text{ELPR}(\boldsymbol \theta_\nv) > c$, then $\Delta(\boldsymbol \theta_\nv) > c$.
```

@@

Additionally, S-unstable RBM model sequences may be formally connected to the near-degeneracy of Section \@ref(near-degeneracy) (in which model sequences place all probability on a small portion of their sample spaces). To see this, define an arbitrary modal set of possible outcomes (i.e. set of highest probability outcomes) in RBM models with parameters $\boldsymbol \theta_\nv, \nv \ge 1$ as
$$
M_{\epsilon, \boldsymbol \theta_\nv} \equiv \left\{\boldsymbol v \in \mathcal{C}^\nv: \log P_{\boldsymbol \theta_\nv}(\boldsymbol v) > (1-\epsilon)\max\limits_{\boldsymbol v^*}P_{\boldsymbol \theta_\nv}(\boldsymbol v^*) + \epsilon\min\limits_{\boldsymbol v^*}P_{\boldsymbol \theta_\nv}(\boldsymbol v^*) \right\}
$$
for a given $0 < \epsilon < 1$. Then S-unstable model sequences are guaranteed to be degenerate, as the following result shows.

@@

In other words, S-unstable RBM model sequences are guaranteed to stack up all probability on a specific set of outcomes for visibles, which could potentially be arbitrarily narrow. Proofs of Propositions \@ref(prp:instab1) and \@ref(prp:degen) appear in @kaplan2017note. These findings also have counterparts in results in @schweinberger2011instability, but unlike results there, we do not limit consideration to exponential family forms with a fixed number of parameters.

@@

Secondly, we can look at $\text{ELPR}(\boldsymbol \theta)$ from (\@ref(eq:elpr)) for this tiny model in order to consider model instability as a function of RBM parameters. Recall that large values of $\text{ELPR}(\boldsymbol \theta)$ are associated with an extreme sensitivity of the model probabilities $f_{\boldsymbol \theta}(\boldsymbol x)$ to small changes in $\boldsymbol x$ (see Proposition \@ref(prp:instab1)). The quantity $\text{ELPR}(\boldsymbol \theta)$ for this small RBM is 
\begin{align*}
\text{ELPR}(\boldsymbol \theta) &= \log \left[\frac{\max\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{\min\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}\right] = \log \left[\frac{\max\limits_{v_1 \in \mathcal{C}}\sum\limits_{h_1 \in \mathcal{C}}\exp\left\{\theta_{11} h_1 v_1 + \theta_{h_1}h_1 + \theta_{v_1} v_1 \right\}}{\min\limits_{v_1 \in \mathcal{C}}\sum\limits_{h_1 \in \mathcal{C}}\exp\left\{\theta_{11} h_1 v_1 + \theta_{h_1}h_1 + \theta_{v_1} v_1 \right\}}\right]. \\
\end{align*}

Figure \@ref(fig:instab) shows contour plots of $\text{ELPR}(\boldsymbol \theta)/\nv$ for various values of $\boldsymbol \theta$ in this model with $\nv = 1$. We can see that this quantity is large for large magnitudes of $\boldsymbol \theta$, especially for large values of the dependence/interaction parameter $\theta_{11}$. This suggests instability as $|\boldsymbol \theta|$ becomes large, agreeing also with the concerns about near-degeneracy produced by consideration of $\boldsymbol \mu(\boldsymbol \theta)$. \par

```{r instab, fig.cap="$\\text{ELPR}(\\boldsymbol \\theta)/\\nv$ for various values of $\\boldsymbol \\theta$ for the tiny example model. Recall here $\\nv$ is the number of visible nodes and here is $1$. This quantity is large for large magnitudes of $\\boldsymbol \\theta$.", fig.height = 4, fig.align='center'}
expand.grid(theta_h1 = seq(-5, 5, length.out = 21),
            theta_v1 = seq(-5, 5, length.out = 21),
            theta_11 = seq(-5, 5, length.out = 5)) %>%
  data.frame() %>%
  group_by(theta_v1, theta_h1, theta_11) %>%
  do(data.frame(Q = cbind(possibles, possibles %*% t(data.matrix(.))))) %>%
  group_by(theta_v1, theta_h1, theta_11, Q.v1) %>%
  do(data.frame(marginal = sum(exp(.$Q.V4)))) %>%
  group_by(theta_v1, theta_h1, theta_11) %>%
  do(data.frame(max_marg = max(.$marginal),
                min_marg = min(.$marginal))) %>%
  mutate(elpr = log(max_marg/min_marg)) %>%
  ungroup() %>%
  mutate(theta_11 = factor(theta_11, labels = paste0("theta[11]==", unique(theta_11)))) %>%
  ggplot() +
  geom_tile(aes(theta_v1, theta_h1, fill = elpr)) +
  geom_contour(aes(theta_v1, theta_h1, z = elpr), colour = "black") +
  facet_grid(. ~ theta_11, labeller = label_parsed) +
  xlab(expression(theta[v1])) +
  ylab(expression(theta[h1])) +
  scale_fill_gradient(expression(frac(ELPR(theta), V)), low = "yellow", high = "red") +
  theme(aspect.ratio = 1)
```

@@

To explore the impact of RBM parameter vector $\boldsymbol \theta$ magnitude on near-degeneracy, instability, and uninterpretability, we consider models of small size. For $\nh, \nv \in \{1, \dots, 4\}$, we sample `r dim(res$samp[[1]])[1]` values of $\boldsymbol \theta$ with various magnitudes (details to follow). For each set of parameters we then calculate metrics of model impropriety introduced in Section \@ref(degeneracy-instability-and-uninterpretability) based on $\boldsymbol \mu(\boldsymbol \theta)$, $\text{ELPR}(\boldsymbol \theta)/\nv$, and the absolute coordinates of $\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] -   \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right]$. In the case of near-degeneracy, we classify each model as near-degenerate based on the distance of $\boldsymbol \mu(\boldsymbol \theta)$ from the boundary of the convex hull of $\mathcal{T}$ and look at the fraction of models that are "near-degenerate," meaning they are within a small distance $\epsilon > 0$ of the boundary of the convex hull. We define "small" through a rough estimation of the volume of the hull for each model size. We pick $\epsilon_0 = 0.05$ for $\nh=\nv=1$ and then, for every other $\nh$ and $\nv$, set $m=\nh+\nv+\nv*\nh$ and pick $\epsilon$ so that $1-(1-2\epsilon_0)^3 = 1 - (1-2\epsilon)^m$. In this way, we roughly scale the volume of the "small distance" to the boundary of the convex hull to be equivalent across model dimensions.

@@

three_ways %>%
  mutate(r1 = round(r1/(H + V), 8), r2 = round(r2/(H * V), 8)) %>% 
  ggplot() +
  geom_tile(aes(x = r1, y = r2, fill = scaled_mean_elpr)) +
  geom_contour(aes(x = r1, y = r2, z = scaled_mean_elpr), colour = "black", bins = 8) +
  geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
  scale_fill_gradient(expression(paste("Mean ", frac(ELPR(theta), V))), low = "yellow", high = "red") +
  facet_grid(Visibles~Hiddens, labeller = label_parsed) +
  xlab(expression(paste(paste("||", theta[main], "||", "/(", n[h]," + ", n[v], ")")))) +
  ylab(expression(paste(paste("||", theta[interaction], "||", "/(", n[h], "*", n[v], ")")))) +
  theme(aspect.ratio = 1) -> p.elpr

@@

```{r instab-plots, fig.cap="The sample mean value of $\\text{ELPR}(\\boldsymbol \\theta)/V$ at each grid point for each combination of magnitude of $\\boldsymbol \\theta$ and model size. As the magnitude of $\\boldsymbol \\theta$ grows, so does the value of this metric, indicating typical instability in the model.", fig.height=3.5, fig.align="ht!"}
p.elpr
```

@@

for the outcomes $(\pm 1, \pm 1)$, the appendix provides technical details to show parameter values $(\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})$ may be chosen to match any given cell probabilities with arbitrary closeness.  In fact, when the cell probabilities (\@ref(eq:0)) are all strictly positive, parameters in the RBM model can be specified to reflect these probabilities exactly. And, when one or more of the cell probabilities (\@ref(eq:0)) are zero, the corresponding RBM probabilities $P(V_1=v_1,V_2=_v2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})$ may never be identically zero (due to exponential terms in the model) but parameters can be still selected to make the appropriate RBM cell probabilities arbitrarily small. Furthermore, as development in the appendix shows, not only is it possible to approximate any distribution on the visibles arbitrarily well [cf. @montufar2011refinements], but dramatically different parameter settings can induce the same RBM model (beyond mere symmetries in parameterization). A further, and somewhat odd consequence of the RBM parameterization is then that, when fitting the RBM model by likelihood-based methods, we already know the nature of the answer before we begin: namely, such  fitting will simply reproduce the empirical distribution of the training data if sufficiently many hiddens are in the model and there can be no model refinements or smoothing from the RBM. That is, based on a random sample of vectors of visible variables, the model for the cell probabilities with the highest likelihood over  *all possible model classes* (i.e., RBM-based or not) is the empirical distribution, and the over-parameterization of the RBM model itself ensures that this empirical distribution can be arbitrarily well-approximated. For illustration, considering the simple example from above with $n$ iid observations, each consisting of two realized visibles $(V_1,V_2)$ the empirical cell frequencies from the sample are the highest likelihood choices for the cell probabilities $p_{(-1,-1)},p_{(1,-1)}, p_{(-1,1)},p_{(1,1)}$ in (\@ref(eq:0)) before any parametric model is introduced for improved analysis, and the discussion above indicates that RBM model parameters can be chosen to re-create this empirical distribution to an arbitrarily close degree. Not only does RBM model fitting based on ML seek to reproduce the empirical distribution, whenever this empirical distribution contains empty cells, fitting steps for the RBM model will further aim to choose parameters that necessarily diverge to infinity in magnitude in order to zero-out the corresponding RBM cell probabilities. In data applications with a large sample space, it is unlikely that the training set will include at least one of each possible vector outcome (unlike this small example). This implies that some RBM model parameters must diverge to $+\infty$ to mimic the empirical distribution with empty cells and, as we have already discussed in Section \@ref(explorations-of-model-properties-through-simulation), large magnitudes of $\boldsymbol \theta$ lead to model impropriety in the RBM.

@@

>>>>

In addition to such degeneracy, we find that RBMs can easily exhibit types of instability, related to how sensitive the model probability structure can be to small changes in data outcomes. In practice, this may be seen when a single pixel change in an image results in a wildly different classification in an image classification problem. Occurrences of such behavior have recently been documented in RBMs [@li2014biclustering], as well as other deep architectures [@szegedy2013intriguing; @nguyen2014deep]. We describe potential issues of model instability, degeneracy and uninterpretability for the RBM class (which are related properties) and examine the presence of these in Section \@ref(explorations-of-model-properties-through-simulation) through simulations of small, manageable examples.

@@

Notions of weight-decay (penalization) and sparsity (regularization) have been suggested in the RBM literature as practical remedies for both over-fitting and poor mixing in the Markov chain during fitting [@hinton2010practical; @tieleman2008training; @cho2012tikhonov]. Both $L_1$ and $L_2$ type penalties are mentioned to achieve weight-decay, though the benefits of these particular forms are unknown in different situations. However, the degree to which regularization and penalization are used by practitioners is not clear because these concepts are not (by default) a part of the accepted fitting methodology [@hinton2002training; @carreira2005contrastive]. In this paper, we attempt to address the concerns with overfitting and poor mixing in an alternative (and perhaps more transparent) manner, via specification of a Bayesian prior in Section \@ref(bayesian-model-fitting).

This paper is structured as follows. Section \@ref(background) formally defines the RBM including the joint distribution of hidden and visible variables and explains the model's connection to deep learning. Additionally, measures of model impropriety and methods of quantifying/detecting it are defined. Section \@ref(explorations-of-model-properties-through-simulation) details our explorations into the model behavior and potential propriety issues with the RBM class. We examine three Bayesian fitting techniques intended to avoid model impropriety issues raised in Section \@ref(model-fitting) and conclude with a discussion in Section \@ref(discussion). Supplementary online material provides proofs for results on RBM parameterizations and data codings described in Section \@ref(data-encoding).

While applications of the RBM have been claimed to produce some successes in classification problems, it is unclear if the model class allows one to go beyond fitting to other statistical matters of importance in using the models, such as quantification of the uncertainty of estimation and the formulation of predictive distributions. These are closely tied to the probability properties of the RBM model for explaining data generation and the current exposition intends to contribute to a better understanding in this direction.

@@

Considering exponential families of distributions, @schweinberger2011instability introduced a concept of model deficiency related to *instability*. Instability can be roughly thought of as excessive sensitivity in the model, where small changes in the components of potential data outcomes, $\boldsymbol x$, may lead to substantial changes in the probability mass function $f_{\boldsymbol \theta}(\boldsymbol x)$. Furthermore, model instability can be viewed on a spectrum of potential sensitivity in probability structure, with model degeneracy included as a extreme or limiting case of instability. To quantify "instability" more rigorously (particularly beyond the definition given by @schweinberger2011instability) it is useful to consider how RBM models might be expanded to incorporate more and more visibles. When increasing the size of RBM models, it becomes necessary to grow the number of model parameters (and in this process one may also arbitrarily expand the number of hidden variables used). To this end, let $\boldsymbol \theta_{\nv} \equiv (\theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}, \theta_{11}, \dots, \theta_{\nv \nh}), \nv \ge 1$, denote an element of a sequence of RBM parameters indexed by the number $\nv$ of visibles $(V_1, \dots, V_\nv)$ and define a log-ratio of extremal
probabilities (LREP) of the RBM model at $\boldsymbol \theta_{\nv}$ as
\begin{align}
\frac{1}{\nv} \log \left[\frac{\max\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{\min\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}\right] \equiv \frac{1}{\nv} \text{LREP}(\boldsymbol \theta_\nv) (\#eq:elpr)
\end{align}
where $P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv) \propto \sum_{(h_1, \dots, h_\nh) \in \mathcal{C}^\nh}\exp\left(\sum_{i = 1}^\nv \theta_{v_i} v_i + \sum_{j = 1}^\nh \theta_{h_j} h_j + \sum_{i = 1}^\nv \sum_{j = 1}^\nh \theta_{ij} v_i h_j\right)$ is the RBM probability of observing outcome $(v_1, \dots, v_\nv)$ for the visible variables $(V_1, \dots, V_\nv)$ under parameter vector $\boldsymbol \theta_\nv$, after marginalization of hidden variables.

In formulating a RBM model for a potentially large number of visibles (i.e., as $\nv \rightarrow \infty$), we will say that the ratio (\@ref(eq:elpr)) needs to stay bounded for a sequence of RBM models to be stable. That is, we make the following convention.
```{definition, name = "S-unstable RBM", echo=TRUE}
Let $\boldsymbol \theta_\nv \in \mathbb{R}^{\nv + \nh +  \nh*\nv}, \nv \ge 1$, be an element of a sequence of RBM parameters where the number of hiddens, $\nh \equiv \nh(\nv) \ge 1$, can be an arbitrary function of the number $\nv$ of visibles. A RBM model formulation is \emph{Schweinberger-unstable} or \emph{S-unstable} if
\begin{align*}
\lim\limits_{\nv \rightarrow \infty} \frac{1}{\nv} \text{LREP}(\boldsymbol \theta_\nv) = \infty.
\end{align*}
```
In other words, a RBM model sequence is unstable if, given any $c > 0$, there exists an integer $n_c > 0$ so that $\frac{1}{\nv}\text{LREP}(\boldsymbol \theta_\nv) > c$ for all $\nv \ge n_c$. This definition of *S-unstable* is a generalization or re-interpretation of the "unstable" concept of @schweinberger2011instability in that here RBM models for visibles $(v_1, \dots, v_\nv)$ do not form an exponential family and the dimensionality of $\boldsymbol \theta_\nv$ is not fixed, but rather grows with $\nv$.

@@

```{proposition, label = "instab1", echo=TRUE}
Let $c > 0$ and let $\text{LREP}(\boldsymbol \theta_\nv)$ be as in (\@ref(eq:elpr)) for an integer $\nv \ge 1$. If $\frac{1}{\nv}\text{LREP}(\boldsymbol \theta_\nv) > c$, then $\Delta(\boldsymbol \theta_\nv) > c$.
```

@@

Additionally, S-unstable RBM model sequences may be formally connected to the near-degeneracy of Section \@ref(near-degeneracy) (in which model sequences place all probability on a small portion of their sample spaces). To see this, define an arbitrary modal set of possible outcomes (i.e. set of highest probability outcomes) in RBM models with parameters $\boldsymbol \theta_\nv, \nv \ge 1$ as
$$
M_{\epsilon, \boldsymbol \theta_\nv} \equiv \left\{\boldsymbol v \in \mathcal{C}^\nv: \log P_{\boldsymbol \theta_\nv}(\boldsymbol v) > (1-\epsilon)\max\log\limits_{\boldsymbol v^*}P_{\boldsymbol \theta_\nv}(\boldsymbol v^*) + \epsilon\min\log\limits_{\boldsymbol v^*}P_{\boldsymbol \theta_\nv}(\boldsymbol v^*) \right\}
$$
for a given $0 < \epsilon < 1$. Then S-unstable model sequences are guaranteed to be degenerate, as the following result shows.

@@

In other words, S-unstable RBM model sequences are guaranteed to stack up all probability on a specific set of outcomes for visibles, which could potentially be arbitrarily narrow. Proofs of Propositions \@ref(prp:instab1) and \@ref(prp:degen) appear in @kaplan2016propriety. These findings also have counterparts in results in @schweinberger2011instability, but unlike results there, we do not limit consideration to exponential family forms with a fixed number of parameters.

@@

Secondly, we can look at $\text{LREP}(\boldsymbol \theta)$ from (\@ref(eq:elpr)) for this tiny model in order to consider model instability as a function of RBM parameters. Recall that large values of $\text{LREP}(\boldsymbol \theta)$ are associated with an extreme sensitivity of the model probabilities $f_{\boldsymbol \theta}(\boldsymbol x)$ to small changes in $\boldsymbol x$ (see Proposition \@ref(prp:instab1)). The quantity $\text{LREP}(\boldsymbol \theta)$ for this small RBM is 
\begin{align*}
\text{LREP}(\boldsymbol \theta) &= \log \left[\frac{\max\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{\min\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}\right] = \log \left[\frac{\max\limits_{v_1 \in \mathcal{C}}\sum\limits_{h_1 \in \mathcal{C}}\exp\left\{\theta_{11} h_1 v_1 + \theta_{h_1}h_1 + \theta_{v_1} v_1 \right\}}{\min\limits_{v_1 \in \mathcal{C}}\sum\limits_{h_1 \in \mathcal{C}}\exp\left\{\theta_{11} h_1 v_1 + \theta_{h_1}h_1 + \theta_{v_1} v_1 \right\}}\right]. \\
\end{align*}

Figure \@ref(fig:instab) shows contour plots of $\text{LREP}(\boldsymbol \theta)/\nv$ for various values of $\boldsymbol \theta$ in this model with $\nv = 1$. We can see that this quantity is large for large magnitudes of $\boldsymbol \theta$, especially for large values of the dependence/interaction parameter $\theta_{11}$. This suggests instability as $|\boldsymbol \theta|$ becomes large, agreeing also with the concerns about near-degeneracy produced by consideration of $\boldsymbol \mu(\boldsymbol \theta)$. \par

```{r instab, fig.cap="$\\text{LREP}(\\boldsymbol \\theta)/\\nv$ for various values of $\\boldsymbol \\theta$ for the tiny example model. Recall here $\\nv$ is the number of visible nodes and here is $1$. This quantity is large for large magnitudes of $\\boldsymbol \\theta$.", fig.height = 4, fig.align='center'}
expand.grid(theta_h1 = seq(-5, 5, length.out = 21),
            theta_v1 = seq(-5, 5, length.out = 21),
            theta_11 = seq(-5, 5, length.out = 5)) %>%
  data.frame() %>%
  group_by(theta_v1, theta_h1, theta_11) %>%
  do(data.frame(Q = cbind(possibles, possibles %*% t(data.matrix(.))))) %>%
  group_by(theta_v1, theta_h1, theta_11, Q.v1) %>%
  do(data.frame(marginal = sum(exp(.$Q.V4)))) %>%
  group_by(theta_v1, theta_h1, theta_11) %>%
  do(data.frame(max_marg = max(.$marginal),
                min_marg = min(.$marginal))) %>%
  mutate(elpr = log(max_marg/min_marg)) %>%
  ungroup() %>%
  mutate(theta_11 = factor(theta_11, labels = paste0("theta[11]==", unique(theta_11)))) %>%
  ggplot() +
  geom_tile(aes(theta_v1, theta_h1, fill = elpr)) +
  geom_contour(aes(theta_v1, theta_h1, z = elpr), colour = "black") +
  facet_grid(. ~ theta_11, labeller = label_parsed) +
  xlab(expression(theta[v1])) +
  ylab(expression(theta[h1])) +
  scale_fill_gradient(expression(frac(LREP(theta), V)), low = "yellow", high = "red") +
  theme(aspect.ratio = 1)
```

@@

To explore the impact of RBM parameter vector $\boldsymbol \theta$ magnitude on near-degeneracy, instability, and uninterpretability, we consider models of small size. For $\nh, \nv \in \{1, \dots, 4\}$, we sample `r dim(res$samp[[1]])[1]` values of $\boldsymbol \theta$ with various magnitudes (details to follow). For each set of parameters we then calculate metrics of model impropriety introduced in Section \@ref(degeneracy-instability-and-uninterpretability) based on $\boldsymbol \mu(\boldsymbol \theta)$, $\text{LREP}(\boldsymbol \theta)/\nv$, and the absolute coordinates of $\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] -   \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right]$. In the case of near-degeneracy, we classify each model as near-degenerate based on the distance of $\boldsymbol \mu(\boldsymbol \theta)$ from the boundary of the convex hull of $\mathcal{T}$ and look at the fraction of models that are "near-degenerate," meaning they are within a small distance $\epsilon > 0$ of the boundary of the convex hull. We define "small" through a rough estimation of the volume of the hull for each model size. We pick $\epsilon_0 = 0.05$ for $\nh=\nv=1$ and then, for every other $\nh$ and $\nv$, set $m=\nh+\nv+\nv*\nh$ and pick $\epsilon$ so that $1-(1-2\epsilon_0)^3 = 1 - (1-2\epsilon)^m$. In this way, we roughly scale the volume of the "small distance" to the boundary of the convex hull to be equivalent across model dimensions.

@@

three_ways %>%
  mutate(r1 = round(r1/(H + V), 8), r2 = round(r2/(H * V), 8)) %>% 
  ggplot() +
  geom_tile(aes(x = r1, y = r2, fill = scaled_mean_elpr)) +
  geom_contour(aes(x = r1, y = r2, z = scaled_mean_elpr), colour = "black", bins = 8) +
  geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
  scale_fill_gradient(expression(paste("Mean ", frac(LREP(theta), V))), low = "yellow", high = "red") +
  facet_grid(Visibles~Hiddens, labeller = label_parsed) +
  xlab(expression(paste(paste("||", theta[main], "||", "/(", n[h]," + ", n[v], ")")))) +
  ylab(expression(paste(paste("||", theta[interaction], "||", "/(", n[h], "*", n[v], ")")))) +
  theme(aspect.ratio = 1) -> p.elpr

@@

```{r instab-plots, fig.cap="The sample mean value of $\\text{LREP}(\\boldsymbol \\theta)/V$ at each grid point for each combination of magnitude of $\\boldsymbol \\theta$ and model size. As the magnitude of $\\boldsymbol \\theta$ grows, so does the value of this metric, indicating typical instability in the model.", fig.height=3.5, fig.align="ht!"}
p.elpr
```

@@

for the outcomes $(\pm 1, \pm 1)$, the online appendix provides technical details to show parameter values $(\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})$ may be chosen to match any given cell probabilities with arbitrary closeness.  In fact, when the cell probabilities (\@ref(eq:0)) are all strictly positive, parameters in the RBM model can be specified to reflect these probabilities exactly. And, when one or more of the cell probabilities (\@ref(eq:0)) are zero, the corresponding RBM probabilities $P(V_1=v_1,V_2=_v2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})$ may never be identically zero (due to exponential terms in the model) but parameters can be still selected to make the appropriate RBM cell probabilities arbitrarily small. Furthermore, as development in the online appendix shows, not only is it possible to approximate any distribution on the visibles arbitrarily well [cf. @montufar2011refinements], but dramatically different parameter settings can induce the same RBM model (beyond mere symmetries in parameterization). A further, and somewhat odd consequence of the RBM parameterization is then that, when fitting the RBM model by likelihood-based methods, we already know the nature of the answer before we begin: namely, such  fitting will simply reproduce the empirical distribution of the training data if sufficiently many hiddens are in the model and there can be no model refinements or smoothing from the RBM. That is, based on a random sample of vectors of visible variables, the model for the cell probabilities with the highest likelihood over  *all possible model classes* (i.e., RBM-based or not) is the empirical distribution, and the over-parameterization of the RBM model itself ensures that this empirical distribution can be arbitrarily well-approximated. For illustration, considering the simple example from above with $n$ iid observations, each consisting of two realized visibles $(V_1,V_2)$ the empirical cell frequencies from the sample are the highest likelihood choices for the cell probabilities $p_{(-1,-1)},p_{(1,-1)}, p_{(-1,1)},p_{(1,1)}$ in (\@ref(eq:0)) before any parametric model is introduced for improved analysis, and the discussion above indicates that RBM model parameters can be chosen to re-create this empirical distribution to an arbitrarily close degree. Not only does RBM model fitting based on ML seek to reproduce the empirical distribution, whenever this empirical distribution contains empty cells, fitting steps for the RBM model will further aim to choose parameters that necessarily diverge to infinity in magnitude in order to zero-out the corresponding RBM cell probabilities. In data applications with a large sample space, it is unlikely that the training set will include at least one of each possible vector outcome (unlike this small example). This implies that some RBM model parameters must diverge to $+\infty$ to mimic the empirical distribution with empty cells and, as we have already discussed in Section \@ref(explorations-of-model-properties-through-simulation), large magnitudes of $\boldsymbol \theta$ lead to model impropriety in the RBM.

@@

# References {-}
